\bf Model Visualization Techniques for a Social Network Model

### 2017-08-28 paper.Rnw
could you add the senate csv files?

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, were designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}. Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
%hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

@@

\hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
\st{XXX see line 1257 for plot. where should I put it? }

@@

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

\subsubsection{Continuous Time Markov Chain}
\st{
In the continuous-time Markov chain literature, see for instance \citet{ctmc}, the chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships because self-ties are not allowed.  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Since only one tie changes at a time in the CTMC, there are only $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these represent the possible states that are one edge different from a given state, and the additional non-zero entry is for the state to remain unchanged. All other entries in a row are zero because those network states cannot be reached from the current state by just one change.

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. Estimating the parameters in these models is difficult, but thanks to the SIENA software, we have an accessible way to fit SAOMs to network observations.

>>>>

Some commonly used layout algorithms, such as  the Kamada-Kawai layout \citep{kamadakawai} and the Fruchterman-Reingold layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. In these algorithms, the edges of the network act as springs pushing and pulling the nodes in a low dimensional (usually two-dimensional) space. Another algorithm uses multi-dimensional scaling, relying on distance metric and computing a matrix whose entries represent the ``distance" between every pair of nodes. There are also layout algorithms that use properties of the adjacency matrix, like its eigenstructure, to place the nodes in 2D space \citep{drawingalgs}. The node-link diagram using the Kamada-Kawai layout algorithm for our toy network is shown in Figure~\ref{fig:toyplots}. Unless otherwise stated, all other node-link diagrams in this paper will use the Kamada-Kawai layout. %Other algorithms use properties of the network's adjaceny matrix in order to place the nodes.
%hh{XXX how does the adjacency matrix feature here? The adjacency matrix is a different mathematical representation of a network, but describes the same network as a node-link diagram.} \st{XXX I'm taking this to mean I should describe adjacency matrices more, so I've added a discussion below. XXX }

@@

\hh{XXX we tried to estimate the rate based on the simulations - you could include a histogram of the number of changes for the 1000 simulations in the viz part XXX}
\st{XXX see line 1257 for plot. where should I put it? }
\hh{XXX Put a reference to the plot here.}

@@

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node. \st{These effects should seem familiar to readers used to working with the classical exponential random graph model (ERGM) for networks. The outdegree represents the propensity of nodes with a lot of outgoing ties to form more outgoing ties (the "rich get richer" effect), and the reciprocity parameter measures the tendency of outgoing ties to be returned within a network. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software that we use to fit the SAOMs, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation depends only on the structure of the network, like the outdegree and reciprocity parameters mentioned above. The parameters are included when the researcher hypothesizes that they will model underlying mechanisms of network change. They hope to answer questions such as, ``How does the existing network structure influence change in the network?" %\hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)} \hh{I guess you addressed this comment very thoroughly :)}
% and ``How do the behavior and characteristics of the nodes influence change in the network?" put this sentence below somewhere
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, with the additional inclusion of node-level covariates of interest. The covariate effects are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

@@

\subsubsection{Continuous Time Markov Chain (CTMC)}
\st{
In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

@@

Thus, the rate of change between any two states, $x^b$ and $x^c$, that differ by only one tie $x_{ij}$, is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$. Estimating the parameters in these models is difficult, but thanks to the SIENA software \citep(Rsiena), we have an accessible way to fit SAOMs to network observations.
